---
title: "Final Assessment. Dataset Analysis"
author: 'Pau Blasco Roca. NIU: 1600959'
date: "28/5/2022"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A statistical approach to the study of mammal sleep patterns in relation to body and brain weight

The intent of this paper is to study the relationship between the
physical qualities of different species of mammals and their respective 
sleep patterns. These studies consist on an initial linear regression 
model, followed by a bootstrap approach. After
that, further correlation and bootstrap studies are performed by
separating mammals in several categories.

```{r include=FALSE}
library(here)
library(readr)
library(ggplot2)
library(tidyverse)
library(plyr)
library(corrplot)
library(cowplot)
library(boot)
```

```{r}
#Initial Linear Regression fittings for some of our data variables.
myData = msleep
myData
par(mfrow=c(1,2))
P1<-ggplot(myData, aes(x=sleep_total, y=sleep_rem)) +
  geom_point() +
  labs(x='Total Sleep Time (h)', y='Total Rem Time (h)') +
  stat_smooth(method='lm', color = "turquoise4")+
  theme(plot.title = element_text(hjust=0.5, size=15, face='bold'), aspect.ratio = 1) 
P2<-ggplot(myData, aes(x=brainwt, y=sleep_cycle)) +
  geom_point() +
  labs(x='Brain Weight (kg)', y='Sleep Cycle Length (h)') +
  ylim(0,2)+
  xlim(0,1.4)+
  stat_smooth(method='lm', color = "turquoise4")+
  theme(plot.title = element_text(hjust=0.5, size=15, face='bold'), aspect.ratio = 1)
P<-plot_grid(P1, P2)
P
linearMod1 <- lm(sleep_total ~ sleep_rem, data=myData)  # build linear regression model on full data
summary(linearMod1)
linearMod2 <- lm(brainwt ~ sleep_cycle, data=myData)  # build linear regression model on full data
summary(linearMod2)
```

```{r}
#Studying only numeric variables, without distinguishing by mammal category
myData = msleep %>% select(sleep_total,sleep_rem,sleep_cycle,bodywt,brainwt)
res <- cor(myData, use = "complete.obs")
round(res, 2)
#we calculate the correlation matrix between all the variables in the dataset
corrplot(res, method="number", type = "lower", order = "original", 
         tl.col = "black", tl.srt = 45)
```


```{r}
#Parametric bootstrap to prove H0="body weight influences total sleep time" (the slope is not 0)
set.seed(1111)
myData = msleep %>% select(bodywt,sleep_total)
N=10000
siz=35
x <- 1:N;
for (i in 1:N){
  boot  = na.omit(myData[sample(na.omit(siz),(siz),replace = TRUE),])
  bootX = as.matrix(boot[,c(2)])
  bootY = as.matrix(boot[,c(1)])
  coefs = lm(bootX ~ bootY)$coefficients
  x[i]= coefs["bootY"]
}

#We find the standard errors without having to derive an equation manually
x = sort(x)
SE_slope = sd(x)
mean_slope = mean(x)
#outputting the estimate and the Standard Error
SE_slope
mean_slope
#Our CI is determined by the 500th and 9500th value of the ordered set of results
p<- ggplot() + aes(x)+ geom_histogram(binwidth=0.0012, colour="black", fill="#30d5c8")  + xlim(-0.025,0.01) +
  ggtitle("Bootstrap with 0.90 CI") + xlab("BDW and TST Correlation Coefficient") + ylab("Frequency") + 
  theme(plot.title = element_text(color="black", size=20, face="bold"), axis.title.x = element_text(color="black", size=14, face="bold"), axis.title.y = element_text(color="black", size=14, face="bold")) +
  geom_vline(xintercept = x[500], color = "blue", size=1.5) + 
  geom_vline(xintercept = x[9500], color = "blue", size=1.5)
p
```

```{r}
#Non parametric bootstrap to prove H0="body weight influences total sleep time" (the slope is not 0)
set.seed(2222)
myData = msleep %>% select(bodywt,sleep_total)
correlator = function(base,i){return(cor(base[i,]$bodywt, base[i,]$sleep_total))}
crtl_boot = boot(myData,correlator,R=10000)
#Our CI is determined following normal distribution quantiles.
boot.ci(boot.out = crtl_boot, type = c("norm"), conf = 0.90)
```

```{r}
#Parametric Bootstrap to prove H0="brain weight influences sleep cycle length" (the slope is not 0)
set.seed(5555)
myData = msleep %>% select(sleep_cycle,brainwt)
N=10000
siz=35
x <- 1:N;
for (i in 1:N){
  boot  = na.omit(myData[sample(na.omit(siz),(siz),replace = TRUE),])
  bootX = as.matrix(boot[,c(2)])
  bootY = as.matrix(boot[,c(1)])
  coefs = lm(bootX ~ bootY)$coefficients
  x[i]= coefs["bootY"]
}

#We find the standard errors without having to derive an equation manually
x = sort(x)
SE_slope = sd(x)
mean_slope = mean(x)
#outputting the estimate and the Standard Error
SE_slope
mean_slope
#Our CI is determined by the 500th and 9500th value of the ordered set of results
p<- ggplot() + aes(x)+ geom_histogram(binwidth=0.05, colour="black", fill="#30d5c8") + 
  geom_vline(xintercept = x[500], color = "blue", size=1.5) + 
  geom_vline(xintercept = x[9500], color = "blue", size=1.5) + 
  ggtitle("Bootstrap with 0.90 CI") + xlab("BRW and SCL Correlation Coefficient") + ylab("Frequency") + 
  theme(plot.title = element_text(color="black", size=20, face="bold"), axis.title.x = element_text(color="black", size=14, face="bold"), axis.title.y = element_text(color="black", size=14, face="bold"))
p
```

```{r}
#Non parametric bootstrap to prove H0="brain weight influences sleep cycle length" (the slope is not 0)
set.seed(2222)
myDataInc = msleep %>% select(sleep_cycle,brainwt)
myData = myDataInc[complete.cases(myDataInc), ]
correlator = function(base,i){return(cor(base[i,]$sleep_cycle, base[i,]$brainwt))}
crtl_boot = boot(myData,correlator,R=10000)
#Our CI is determined following normal distribution quantiles.
boot.ci(boot.out = crtl_boot, type = c("norm"), conf=0.90)
```



```{r}
#brief correlation study dividing the dataset by the mamals' diet.
#NOTE: this does NOT appear in the paper, but it was useful to the investigation.
myDataCarni = msleep %>% filter(vore=="carni") %>% select(sleep_total,sleep_rem,sleep_cycle,bodywt,brainwt)
myDataHerbi = msleep %>% filter(vore=="herbi") %>% select(sleep_total,sleep_rem,sleep_cycle,bodywt,brainwt) 
myDataOmniv = msleep %>% filter(vore=="omni")  %>% select(sleep_total,sleep_rem,sleep_cycle,bodywt,brainwt)

par(mfrow=c(1,3))
res <- cor(myDataCarni, use = "complete.obs")
round(res, 2)
corrplot(res, method="number", type = "lower", order = "original", 
         tl.col = "black", tl.srt = 45)
res <- cor(myDataHerbi, use = "complete.obs")
round(res, 2)
corrplot(res, method="number", type = "lower", order = "original", 
         tl.col = "black", tl.srt = 45)
res <- cor(myDataOmniv, use = "complete.obs")
round(res, 2)
corrplot(res, method="number", type = "lower", order = "original", 
         tl.col = "black", tl.srt = 45)
```

```{r}
#Pie plot of the distribution of the dataset based on the mammals' diet
data <- data.frame(
  Category=c("Carnivores", "Omnivores", "Herbivores", "Insectivores", "Unclassified"),
  count=c(19, 20, 32, 5, 7)
)
ggplot(data, aes(x="", y=count, fill=reorder(Category,count))) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  #good looking color palette
  scale_fill_manual(values=c("#90E6FC", "#7CD7F7", "#8BCBFC", "#83BDF7", "#84ACFA")) + 
  theme_void() # remove background, grid, numeric labels
```

```{r}
#Data filtering based on their diet
myDataCH = msleep %>% filter(vore=="carni" | vore=="herbi")
par(mfrow=c(1,2))
#Two plots with their respective regression lines
P1<-ggplot(myDataCH, aes(x=sleep_total, y=sleep_cycle, col=vore)) +
  geom_point() +
  xlim(4,14)+
  labs(x='Total Sleep Time (h)', y='Sleep Cycle Length (h)', col='Diet') +
  stat_smooth(method='lm')+
  theme(plot.title = element_text(hjust=0.5, size=15, face='bold'), aspect.ratio = 1.2) 
P2<-ggplot(myDataCH, aes(x=brainwt, y=sleep_cycle, col=vore)) +
  geom_point() +
  xlim(0,0.7)+
  labs(x='Brain Weight (kg)', y='Sleep Cycle Length (h)', col='Diet') +
  stat_smooth(method='lm')+
  theme(plot.title = element_text(hjust=0.5, size=15, face='bold'), aspect.ratio = 1.2)
P<-plot_grid(P1, P2)
P
```



