---
title: "Homework 3 - Covid cases over time"
author: 'Pau Blasco Roca. NIU: 1600959'
date: "15/5/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
#tinytex::install_tinytex()
knitr::opts_chunk$set(echo = TRUE)
```

## Covid cases evolution analysis

The number of registered cases of COVID-19 in Catalonia, day by day, from 27/02/2020 to 31/03/2020:

*y = c(2, 3, 5, 6, 15, 15, 15, 24, 24, 24, 49, 75, 124, 156, 260, 316, 509, 715, 903, 1394, 1866, 2702, 3270, 4203, 4704, 5925, 7864, 9937, 11592, 12940, 1423, 15026, 16157, 18773)*

We want to analyze the evolution of the number of affected individuals as a function of time A(t).

### Exercise 1: Exponential fit

We will fit the data using these two exponential functions.

$A_1(t)=\exp(\beta_0+\beta_1t),\quad A_2(t)=\exp(\beta_0+\beta_1t+\beta_2t^2)$

We are going to assume our data is Poisson distributed. Let's first store our data in a dataset.
```{r}
cases = c(2, 3, 5, 6, 15, 15, 15, 24, 24, 24, 49, 75, 124, 156, 260, 316, 509, 715, 903, 1394, 1866, 2702, 3270, 4203, 4704, 5925, 7864, 9937, 11592, 12940, 1423, 15026, 16157, 18773)
days = seq(1,length(cases),1)
myData <- data.frame("cases" = cases, "days" = days)
```

We will fit the model with both functions and compare their results. We will start using the **glm** method.
```{r}
t <- seq(1, length(myData$days), 1)
t2 <- t*t
f2 <- glm(cases ~ t, family="poisson", data=myData)
summary(f2)$aic
f4 <- glm(cases ~ t + t2, family="poisson",data=myData)
summary(f4)$aic
```

Our results are AIC=24278.81 for the first function and AIC=17978.65 for the second function. We see that, as the second one has a lower value, is the best option for us. Let's plot both functions and study how well they adjust to the data.

```{r echo=FALSE}
b0 <- summary(f2)$coefficients[1]
b1 <- summary(f2)$coefficients[2]
f2c <- function(t) { return (exp(b0 + b1*t)) }
plot(myData$days,  myData$cases, type="p", col="red", ylab="Covid19 cases", xlab="t+n days since Feb 27th 2020" )
par(new=TRUE)
plot(myData$days, f2c(myData$days), type="l", col="green", axes=FALSE, main="Covid cases and glm function A1",xlab="", ylab="" )
```

```{r echo=FALSE}
b0 <- summary(f4)$coefficients[1]
b1 <- summary(f4)$coefficients[2]
b2 <- summary(f4)$coefficients[3]
f4c <- function(t) { return (exp(b0 + b1*t + b2*t**2)) }
plot(myData$days,  myData$cases, type="p", col="red", ylab="Covid19 cases", xlab="t+n days since Feb 27th 2020" )
par(new=TRUE)
plot(myData$days, f4c(myData$days), type="l", col="green", axes=FALSE, main="Covid cases and glm function A2",xlab="", ylab="" )
```

After plotting both options, we can see how the second one adjusts slightly better to our data.

We will now study the model using the **nlm** method. For that, we will first need to compute the maximum likelihood estimator for each function. Luckily, R makes that process really simple.

```{r}
loglik=function(par){
  f = (exp(b0 + b1*t))
  return(-sum(myData$cases*log(f)-f))
}
#These initial parameters seem to give the best results
param=nlm(loglik, p=c(1,0.17),hessian = TRUE)
b0 = param$estimate[1]
b1 = param$estimate[2]
f2c <- function(t) { return (exp(b0 + b1*t)) }
plot(myData$days,  myData$cases, type="p", col="red", ylab="Covid19 cases", xlab="t+n days since Feb 27th 2020" )
par(new=TRUE)
plot(myData$days, f2c(myData$days), type="l", col="green", axes=FALSE, main="Covid cases and nlm function A1",xlab="", ylab="" )
```

```{r}
loglik=function(par){
  f = (exp(b0 + b1*t + b2*t**2))
  return(-sum(myData$cases*log(f)-f))
}
#These initial parameters seem to give the best results
param=nlm(loglik, p=c(2,0.75,-0.01),hessian = TRUE)
b0 = param$estimate[1]
b1 = param$estimate[2]
b2 = param$estimate[3]
f4c <- function(t) { return (exp(b0 + b1*t + b2*t**2)) }
plot(myData$days,  myData$cases, type="p", col="red", ylab="Covid19 cases", xlab="t+n days since Feb 27th 2020" )
par(new=TRUE)
plot(myData$days, f4c(myData$days), type="l", col="green", axes=FALSE, main="Covid cases and nlm function A2",xlab="", ylab="" )
```

Now we must find the expected time at where the cases will stop growing. For that, we will just optimize $A_2$.

```{r}
f4 <- glm(cases ~ t + t2, family="poisson",data=myData)
b0 <- summary(f4)$coefficients[1]
b1 <- summary(f4)$coefficients[2]
b2 <- summary(f4)$coefficients[3]
f4c <- function(t) { return (exp(b0 + b1*t + b2*t**2)) }
optimize(f4c, c(1, length(cases)+10), maximum=TRUE)
plot(seq(1,length(cases)+10,1), f4c(seq(1,length(cases)+10,1)), xlab="t+n days", ylab="expected cases")
```

As we can clearly see, the maximum takes place in day 38.35 (between days 38 and 39). I have plotted the function for 10 more days to make it clear that the maximum takes place in that day.

### Exercise 2: Two-parameters Sigmoid fit

We will fit the data using this sigmoid-like function.

$A(t)=\frac{A_0C}{A_0+(C-A_0)\exp(-\beta t)},\quad\textrm{where }A(0)=A_0\textrm{ and }\beta>0.$

This model is not a **glm** model, as the function is not linear and therefore, glm won't be able to handle it. For this, we need to use **nlm**.

```{r warning=FALSE}
A0 = myData$cases[1]
loglik=function(par){
  f = ((A0*par[1])/(A0+(par[1]-A0)*exp(-par[2]*myData$days)))
  return(-sum(myData$cases*log(f)-f))
}
#We use the initial parameters as the first value of the sample.
param=nlm(loglik, p=c(2,1),hessian = TRUE)
C = param$estimate[1]
b1 = param$estimate[2]
f6 <- function(a0,x){ return((a0*C)/(a0+(C-a0)*exp(-b1*x)))}

plot(myData$days, myData$cases, pch=1, xlab="t+n days", ylab="cases", main="Covid cases and sigmoid fitted function")
lines(myData$days, f6(A0,myData$days), col="green")
```

We can easily compute the limit of this function by studying it mathematically. Our function, fitted to our data, is the following:
$$A(t)=\frac{2·16888}{2+16888\exp(-0.32698 t)},\quad\textrm{where }A(0)=A_0\textrm{ and }\beta>0.$$
Which clearly tends to $C$, in our case $16888$, as time tends to infinity.

To estimate C with a 95% confidence interval, we need to use the following formula:
$$C=\hat{C}\pm1.96\frac{1}{\sqrt{J(\hat{C})}}$$
Where $J(\hat{C})=-l''(\hat{C}))$. Differentiating two times the log-likelihood, we get the hessian matrix:
```{r}
fi= param$hessian
sol <- solve(fi)
sigm <- sol[1,1]
sqrt(sigm)
```

Then, we get a window of $\pm1.96·101.9$ which gives us the following CI: $(16688.276,17087.724)$.